Future challenges for Research Software Engineering

### 2025-09-25 future_challenges_for_RSEng.tex

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.

@@

\subsection{Transformations in Academic Institutions}
\begin{whatis}{}{academic}
Research software is a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
Replacement of staff reaching retirement, combined with the intensified research interest in RSE, has led to the creation of more RSE professorships.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of societal generations.
While sounding negative, this can also be an opportunity that should be used by the RSE community.
If in the time-frame of the next decade, a large number of permanent staff reach retirement and
can be replaced by a new generation of researchers and technical professionals who have seen the
digital transformation first-hand and maybe have actively participated in it through various grassroots initiatives.
This generational change opens up the opportunity for an active change within the research landscape.
A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
A second more progressive step would be to actually appoint new cross-faculty professorships.
On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
Some specialisation trends have already been observed in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated}.
Taking these trends together the generational change should be used as an opportunity to reshape organisational structures to better reflect the work people in research
are actually doing. The RSE community should take an active role in shaping that discussion in this direction till 2035.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

>>>>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software
was only published a couple of years ago~\cite{Anzt2021},
while the foundational competencies of an RSE have only recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.
Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
At the same time, we recognise, the wider international growth of RSEng as a very important development
for people undertaking software development within the research community.
Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
investigation among RSEng communities in other countries.
While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
the wider international community where we feel this is relevant.

@@

\subsection{Transformations in Academic Institutions}
\begin{whatis}{}{academic}
Research software is a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
Some specialisation trends have already been observed in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum
